---
title: "Animated Bar Chart"
author: "ML"
date: "5/3/2020"
output: html_document
---

Load needed packages

```{r}
# check to see if packages are installed. Install them if they are not, then load them into the R session.

ipak <- function(pkg) {
  new.pkg <- pkg[!(pkg %in% installed.packages()[, "Package"])]
  if (length(new.pkg))
    install.packages(new.pkg, dependencies = TRUE)
  sapply(pkg, require, character.only = TRUE)
}

packages <- c("tidyverse", "ggplot2", "gganimate", "png")
ipak(packages)

#install.packages("dplyr") run in case of "there is no package called ‘rlang’" error

```

Prepare analysis files

```{r}
setwd("C:/Users/MarkLaVenia/GitHub/rockr")

twitter_poll_data <-
  read_csv("inst/extdata/raw_twitter_poll_data.csv")

precision_poll_data <- twitter_poll_data %>%
  mutate(album_votes = round(poll_votes * (poll_percent / 100), 0)) %>%
  mutate(poll_percent = (album_votes / poll_votes) * 100) 

rm(twitter_poll_data)

album_finalpolls <-
  precision_poll_data[precision_poll_data$valid %in%  # include only valid polls
              c(1) &
              precision_poll_data$poll_type %in%  # include only final polls
              c("final"),]

save(precision_poll_data, file = "data/precision_poll_data.rda")
rm(precision_poll_data)

band_finalpolls <- album_finalpolls %>%
  group_by(band_name, album_year) %>%
  summarise_at(c("album_votes", "poll_percent"), sum) %>%
  arrange(album_year, band_name)

rm(album_finalpolls)

band_finalpolls_wide <- pivot_wider(
  data = band_finalpolls,
  id_cols = band_name,
  names_from = album_year,
  values_from = c("album_votes", "poll_percent"))

rm(band_finalpolls)

band_finalpolls_long <- band_finalpolls_wide %>%
  pivot_longer(album_votes_1970:poll_percent_1989,
    names_to = c(".value", "year"),
    names_pattern = "(.*)_(.*)",
    values_to = c("var1, var2")) %>%
  arrange(band_name, year) %>%
  mutate(poll_percent_cumulative = cummean(replace_na(poll_percent, 0))) %>%
  mutate(album_votes_cumulative = cumsum(replace_na(album_votes, 0)))

rm(band_finalpolls_wide)

finalpolls_percent_analysisfile <- band_finalpolls_long  %>%
  select(band_name, year, poll_percent_cumulative) %>%
  rename (value = poll_percent_cumulative)

save(finalpolls_percent_analysisfile, file = "data/finalpolls_percent_analysisfile.rda")
rm(finalpolls_percent_analysisfile)

finalpolls_sum_analysisfile <- band_finalpolls_long  %>%
  select(band_name, year, album_votes_cumulative) %>%
  rename (value = album_votes_cumulative)

save(finalpolls_sum_analysisfile, file = "data/finalpolls_sum_analysisfile.rda")
rm(finalpolls_sum_analysisfile)

rm(band_finalpolls_long)

```


```{r}

load("data/precision_poll_data.rda")

album_allpolls <-
  precision_poll_data[precision_poll_data$valid %in%  # include only valid polls
              c(1),]

rm(precision_poll_data)

band_allpolls <- album_allpolls %>%
  group_by(band_name, album_year) %>%
  summarise_at(c("album_votes"), sum) %>%
  arrange(album_year, band_name)

band_allpolls_wide <- pivot_wider(
  data = band_allpolls,
  id_cols = band_name,
  names_from = album_year,
  values_from = c("album_votes"))

rm(band_allpolls)

band_allpolls_long <- band_allpolls_wide %>%
  gather(year, album_votes, 2:22) %>%
  arrange(band_name, year) %>%
  mutate(album_votes_cumulative = cumsum(replace_na(album_votes, 0)))

rm(band_allpolls_wide)

allpolls_sum_analysisfile <- band_allpolls_long  %>%
  select(band_name, year, album_votes_cumulative) %>%
  rename (value = album_votes_cumulative)

save(allpolls_sum_analysisfile, file = "data/allpolls_sum_analysisfile.rda")
rm(allpolls_sum_analysisfile)

rm(band_allpolls_long)

```

#Data Manipulation:
#In this step, We’re going to filter our dataset to retain only the top 10 countries for every given year. We’ll #also create a few more columns that will help us display labels in the plot.

```{r}
load("data/finalpolls_percent_analysisfile.rda")
analysisfile <- finalpolls_percent_analysisfile

#load("data/finalpolls_sum_analysisfile.rda")
#analysisfile <- finalpolls_sum_analysisfile

#load("data/allpolls_sum_analysisfile.rda")
#analysisfile <- allpolls_sum_analysisfile

analysisfile_formatted <- analysisfile %>%
  group_by(year) %>%
  mutate(
    rank = rank(-value),
    value_rel = value / value[rank == 1],
    value_lbl = paste0(" ", round(value / 1))) %>%
  group_by(band_name) %>%
  filter(rank <= 10) %>%
  ungroup()

rm(finalpolls_percent_analysisfile)
rm(analysisfile)

```

#Building Static Plots
#Now that our data is ready to plotted, We’ll build all the required static plots. As you might have seen in the #animation at the top of this post, We’re going to see how the Top 10 Countries based on GDP has changed over the #years in the given dataset. For that we need to build individual plots for each year.

```{r}

#bands <- unique(analysisfile_formatted$band_name)
#bands

mycolors_scale <- c(
  'Anthrax' = '#362f78',
  'Bathory' = '#936663',
  'Black Sabbath' = '#000000',
  'Budgie' = '#8B8B00',
  'Celtic Frost' = '#03A89E',
  'Death' = '#8B1A1A',
  'Deep Purple' = '#551A8B',
  "Guns n' Roses" = '#FFC125',
  'Iron Maiden' = '#458B00',
  'Judas Priest' = '#6E8243',
  'Led Zeppelin' = '#8A4C19',
  'Megadeth' = '#b91e45',
  'Mercyful Fate' = '#FF4500',
  'Metallica' = '#0000CD',
  'Motorhead' = '#2E382F',
  'Night Sun' = '#FCA811',
  'Queen' = '#cb54d6',
  'Rainbow' = '#9932CC',
  'Rush' = '#573420',
  'Scorpions' = '#D25117',
  'Sepultura' = '#C7B26F',
  'Slayer' = '#b10708',
  'Uriah Heep' = '#5c2c2d',
  'Van Halen' = '#3e45c4',
  'Voivod' = '#9F79EE'
)

mycolors_fill <- c(
  'Anthrax' = '#362f78',
  'Bathory' = '#936663',
  'Black Sabbath' = '#000000',
  'Budgie' = '#8B8B00',
  'Celtic Frost' = '#03A89E',
  'Death' = '#8B1A1A',
  'Deep Purple' = '#551A8B',
  "Guns n' Roses" = '#FFC125',
  'Iron Maiden' = '#458B00',
  'Judas Priest' = '#6E8243',
  'Led Zeppelin' = '#8A4C19',
  'Megadeth' = '#b91e45',
  'Mercyful Fate' = '#FF4500',
  'Metallica' = '#0000CD',
  'Motorhead' = '#2E382F',
  'Night Sun' = '#FCA811',
  'Queen' = '#cb54d6',
  'Rainbow' = '#9932CC',
  'Rush' = '#573420',
  'Scorpions' = '#D25117',
  'Sepultura' = '#C7B26F',
  'Slayer' = '#b10708',
  'Uriah Heep' = '#5c2c2d',
  'Van Halen' = '#3e45c4',
  'Voivod' = '#9F79EE'
)

staticplot = ggplot(analysisfile_formatted, aes(rank, group = band_name, fill = as.factor(band_name), color = as.factor(band_name))) +
  geom_tile(aes(y = value / 2, height = value, width = 0.9 ), alpha = 0.8, color = NA) +
#  geom_text(aes(y = 0, label = paste(band_name, " ")), size = 10, vjust = 0.2, hjust = 1) +
  geom_text(aes(y = value, label = paste(value_lbl, "%")), size = 7, hjust = 0) + # use for plot with percentage values
  geom_text(aes(y=value,label = value_lbl), size = 7, hjust = 0) + # use for plots with sum values
  coord_flip(clip = "off", expand = FALSE) +
  scale_y_continuous(labels = scales::comma) +
  scale_x_reverse() +
  scale_colour_manual(values = mycolors_scale) + # to call custom colors
  scale_fill_manual(values = mycolors_fill) + # to call custom colors
  guides(color = FALSE, fill = FALSE) +
  theme(
    axis.line = element_blank(),
    axis.text.x = element_blank(),
    axis.text.y = element_blank(),
    axis.ticks = element_blank(),
    axis.title.x = element_blank(),
    axis.title.y = element_blank(),
    legend.position = "none",
    panel.background = element_blank(),
    panel.border = element_blank(),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    plot.title = element_text(size = 28, hjust = 0.5, face = "bold", colour = "grey", vjust = 1.5),
    plot.subtitle = element_text(size = 22, hjust = 0.5, face = "italic", color = "grey"),
    plot.caption = element_text(size = 19, hjust = 0.5, face = "italic",  color = "grey"),
    plot.background = element_blank(),
    plot.margin = margin(4, 1.8, 1, 7.5, "cm")
  )

staticplot

```

#We will not get into the details of how to static plots are built as that’s pretty much similar to how any #normal plot is built using ggplot2. As you can see in the above code, There are a few key aspects with the #theme() function that are done to make it go well with the animation, like – Only Vertical Grid Lines are drawn #and Legends, Axes Title and few more components are removed from the plot.

#Animation
#The key function here is transition_states() which stitches the individual static plots together by year. #view_follow() is used to give a view as if the background lines (gridlines) are moving as the animation is #progressing.


```{r}
anim = staticplot + transition_states(year, transition_length = 4, state_length = 1) +
  view_follow(fixed_x = TRUE)  +
  
  labs(title = 'Best album cumulative percentage of votes aggregated by band: {closest_state}', # for final polls percentages
       
#  labs(title = 'Best album cumulative votes aggregated by band: {closest_state}', # for final and all polls sums

       subtitle  =  "Top 10 Bands | 1970–1990",
       
       caption  = "Data Source: @nickmoberly Twitter best album of the year final polls (excluding data for bonus and qualifying polls).\nValues indicate the percentage of votes from the polls for 1970 through a given year, weighted equally across years.") # for final polls percentages

#       caption  = "Data Source: @nickmoberly Twitter best album of the year final polls (excluding data for bonus and qualifying polls).\nValues indicate the sum of votes from the polls for 1970 through a given year.") # for final polls sums

#       caption  = "Data Source: @nickmoberly Twitter best album of the year polls, inclusive of all polls (bonus, qualifying, and grand finales).\nValues indicate the sum of votes from the polls for 1970 through a given year.") # for all polls sums

anim 

```

#Rendering
#With the animation being built (ready) and saved in the object anim , It’s time for us to render the animation #using animate() function. The renderer used in the animate() differs based on the type of output file required.
#For GIF File Format:

```{r}
animate(anim, 200, fps = 10, duration = 40, width = 1200, height = 1000,
  
  renderer = gifski_renderer("plots/album_poll_final_percentage.gif")) # for final polls percentages

#  renderer = gifski_renderer("plots/album_poll_final_sum.gif")) # for final polls sums

#  renderer = gifski_renderer("plots/album_poll_all_sum.gif")) # for all polls sums
```

